%*******************************************************
% Acknowledgments
%*******************************************************
\pdfbookmark[1]{Acknowledgements}{acknowledgements}

\bigskip

\begingroup
\let\clearpage\relax
\let\cleardoublepage\relax
\let\cleardoublepage\relax
\chapter*{Acknowledgements}

\def\thanks#1{%
\begingroup
\leftskip1em
\noindent #1
\par
\endgroup
}

First and foremost, I would like to express my sincere gratitude to my advisor Andreas Krause for his excellent mentorship, continued support, for innovative ideas, precise remarks, and critical thinking.
Thank you in particular for your trust and for finding the right balance between guidance and giving me the freedom to pursue my research interests.
Your remarkable insights into which ideas are likely to succeed as well as how to communicate a scientific result never cease to amaze me and will remain with me as a precious lesson for the future.

I would like to extend my gratitude to Marco Cuturi for being my co-advisor from the very beginning. 
Without your invaluable mentorship, this thesis would certainly not be the same. Thank you for empowering me to acquire novel skills and develop and realize my own research agenda.
Also, thank you for providing me with countless opportunities ranging from internships at Google Research and Apple to co-presenting the tutorial at ICML 2023 with me. \\
This combination of advisors has been key for my doctoral studies. \\

Further, I would also like to thank Jure Leskovec for participating in my thesis committee and for providing me with insightful feedback. 
Throughout my academic path, various scientists have left a mark on me as a researcher, chief among them Aviv Regev. While our academic collaboration is still at the very beginning, every interaction with you is intellectually rewarding.

\looseness -1 The thesis would not be the same without the longstanding collaboration with Lucas Pelkmans and Gabriele Gut. First, thank you, Lucas, for serving on my thesis committee.
I have particularly enjoyed our academic ventures at the intersection  of molecular biology and machine learning. Working with you has taught me to think differently about problems and ask different questions, and it has had a lasting effect on this thesis.

I am very grateful to Anne Carpenter and Shantanu Singh for hosting me at the Broad Institute of MIT and Harvard and for the many research discussions we had. I will forever cherish my time in your lab.

I am profoundly indebted to all co-authors and fantastic collaborators with whom I worked on many projects throughout the past years, in particular, Gunnar R\"atsch, Stefan Stark, Ya-Ping Hsieh,  Matteo Pariset, Valentin De Bortoli, Octavian Ganea, Vignesh Ram Somnath, Frederike L{\"u}beck, Laetitia Meng-Papaxanthos, Mojm{\'i}r Mutn{\'y}, Regina Barzilay, Tommi Jaakkola, Connor Coley, and Philippe Schwaller. Thank you especially to Ya-Ping Hsieh and Stefan Stark for hour-long discussions on various projects. 

Further, I am very grateful to my first academic mentors, Stefanie Jegelka, David Alvarez Melis, Roland Eils, Thomas H{\"o}fer, and Lisa Buchauer for guiding me through my first actual research projects.
Thank you, Stefanie, for being a mentor much beyond my time at MIT and for your continued advice. 
Thank you, David, for introducing me to optimal transport many years back, for every coffee during the frequent visits to Cambridge, and in particular, for navigating me through my faculty search and life in academia. I learned so much from you and will be forever grateful.

Lastly, I would not have started my scientific journey without the Life-Science Lab at the German Cancer Research Center. It is what sparked my interest in biology and engineering and set the course for what is here today.
Participating in the "world championship in synthetic biology" was an eye-opener for the power and potential of interdisciplinary research.

\looseness -1 Thanks to all members of the Learning and Adaptive Systems group at ETH Zurich and the ETH AI Center for creating an excellent research environment. Thanks to Lars Lorch, Mohammad Reza Karimi, and Lenart Treven for lengthy discussions on causal inference, sampling, or control theory. 
Thank you, Rita Klute, for the support, for entangling the jungle of bureaucracy, and for overcoming \emph{any} administrative hurdle. \\

I am very grateful to my parents Nele and Egon and my siblings Kaspar, Henriette, and Frieder for their endless love, advice, encouragement, and support.
Growing up in this environment of Freigeister has made me the person I am today.

Thank you to my friends around the globe for making life joyful and adventurous, every party fun, and the difficult moments bearable.

Lastly, thank you to Pol for expanding my world into unknown dimensions: "Jeder Zustand, ja jeder Augenblick [mit dir] ist von unendlichem Wert, denn er ist der Repr{\"a}sentant einer ganzen Ewigkeit."\footnote{Johann Peter Eckermann and Johann Wolfgang von Goethe. Gespr{\"a}che mit Goethe in den letzten Jahren seines Lebens. 1823-1832: 2. Vol. 2. FA Brockhaus, 1836.}

\endgroup