
\chapter*{Notation}

\vskip -2em
\begin{tabularx}{\textwidth}{lX}
 	$\Sigma_n$ & probability simplex of size $n$. \\
 	$(\mu, \nu)$ & measures defined on spaces $(\gX, \gY)$. \\
 	$(u, v)$ & histograms in the simplices $\Sigma_n \times \Sigma_m$. \\
 	$p_\mu = \frac{d\mu}{dx}$ & density with respect to the Lebesgues measure. \\
 	\makecell{$(\mu=\sum_i u_i \delta_{x_i},$ \\ $\,\nu=\sum_j v_j \delta_{y_j})$} & discrete measures defined on spaces $(x_1, \dots, x_n \in \gX, y_1, \dots, y_m \in \gY)$. \\
 	$\mathbf{1}$ & matrix of $\mathbb{R}^{n \times m}$ with all entries identically set to 1. \\
 	$\mathds{1}$ & indicator function. \\
 	$\operatorname{Id}$ & identity map. \\
 	$c(x, y)$ & ground cost, with associated pairwise cost matrix $(c(x_i,y_j))_{ij}$ evaluated on the support of $\mu, \nu$. \\
 	$\|\cdot\|^2_2$ & squared Euclidean norm. \\
 	$\sharp$ & pushforward operator. \\
 	$T: \gX \times \gY$ & Monge map, typically such that $T_\sharp \mu = \nu$. \\
 	$\pi$ & coupling measure between $\mu$ and $\nu$, for discrete measures $\pi = \sum_{ij} \mathbf{P}_{ij} \delta_{(x_i, y_j)}$. \\
 	$\Pi(\mu, \nu)$ & set of couplings, for discrete measures $U(u, v)$. \\
 	$\supp(\pi)$ & support of $\pi$. \\
 	$W(\mu, \nu)$ & Wasserstein distance between measures $\mu$ and $\nu$. \\
 	$H(\pi)$ & entropy of coupling $\pi$. \\
 	$\varepsilon$ & regularization strength of the entropy regularization. \\
 	$(f, g)$ & dual potentials. \\
 	$f^*$ & Legendre transform of function $f$. \\
 	$f^\star$ & optimum of function $f$. \\ 
 	$\varphi$ & convex potential. \\
 	$(\mu_t)_{t=0}^T$ & dynamic measures with $\mu_{t=0} = \mu_0$ and $\mu_{t=T} = \mu_T$. \\
 	$v$ & speed in the dynamic Benamou-Brenier or control in the stochastic optimal control formulation. \\
 	$\Delta$ & Laplace operator. \\
 	$\tau$ & step size. \\
 	$\langle\cdot, \cdot\rangle$ & Euclidean dot-product between vectors. \\
 	$D_{\mathrm{H}}$ & Hellinger distance. \\
 	$\KL$ & Kullback-Leibler divergence. \\
 	$\ell_1$ & Manhattan distance. \\
 	$\ell_2$ & Euclidean distance. \\
 	$\circ$ & Hadamard element-wise product. \\
 	$\sigma$ & noise level. \\
 	$\mathbb{P}_t$ & stochastic process with $t \in [0,1]$. \\
 	$\mathbb{W}_t$ & standard Wiener process. \\
 	$\mathbb{Q}_t$ & reference process, e.g., Brownian motion. \\
 	$Z_t, \hat{Z}_t$ & time-indexed smooth vector fields indicating the forward and backward policy. \\
 	$\gN(\mu, \Sigma)$ & multivariate Gaussian distribution with mean vector $\mu$ and covariance matrix $\Sigma$. \\
 	$\dot{x}$ & first derivative of $x$ taken with respect to time. \\
 	$\ddot{x}$ & second derivative of $x$ taken with respect to time. \\
 	$\theta$ and $\phi$ & parameters of neural networks. \\
 	$\gamma$ & learning rate. \\
 	$\ell$ & loss function. \\
 	$\omega$ & kernel function. \\
 	
\end{tabularx}


\newacronym{OT}{OT}{optimal transport}
\newacronym{SB}{SB}{Schr\"odinger bridge}
\newacronym{DSB}{DSB}{diffusion Schr\"odinger bridge}
\newacronym{GSB}{GSB}{Gaussian Schr\"odinger bridge}
\newacronym{IPF}{IPF}{iterative proportional fitting}
\newacronym{ODE}{ODE}{ordinary differential equation}
\newacronym{SDE}{SDE}{stochastic differential equation}
\newacronym{PDE}{PDE}{partial differential equation}
\newacronym{JKO}{JKO}{Jordan-Kinderlehrer-Otto}
\newacronym{KKT}{KKT}{Karush-Kuhn-Tucker}
\newacronym{PCA}{PCA}{principal component analysis}
\newacronym{iid}{i.i.d.}{independent and identically distributed}
\newacronym{oos}{o.o.s.}{out-of-sample}
\newacronym{ood}{o.o.d.}{out-of-distribution}
\newacronym{lsc}{l.s.c.}{lower semi-continuous}
\newacronym{GAN}{GAN}{generative adversarial network}
\newacronym{NN}{NN}{neural network}
\newacronym{CNN}{CNN}{convolutional neural network}
\newacronym{RNN}{RNN}{recurrent neural network}
\newacronym{GNN}{GNN}{graph neural network}
\newacronym{kNN}{$k$-NN}{$k$-nearest neighbor}
\newacronym{ICNN}{ICNN}{input convex neural network}
\newacronym{PICNN}{PICNN}{partially input convex neural network}
\newacronym{NF}{NF}{normalizing flows}
\newacronym{MLP}{MLP}{multi-layer perceptron}
\newacronym{MMD}{MMD}{maximum-mean-discrepancy}
\newacronym{RMSD}{RMSD}{root-mean-square deviation}
\newacronym{PS}{PS}{perturbation signatures}
\newacronym{SGM}{SGM}{score-based generative model}
\newacronym{SMLD}{SMLD}{score matching with Langevin dynamics}
\newacronym{DDPM}{DDPM}{denoising diffusion probabilistic model}
\newacronym{OU}{OU}{Ornstein-Uhlenbeck}
\newacronym{BDT}{BDT}{Black-Derman-Toy}
\newacronym{BM}{BM}{Brownian motion}
\newacronym{VESDE}{VESDE}{variance exploding \acrshort{SDE}}
\newacronym{VPSDE}{VPSDE}{variance preserving \acrshort{SDE}}
\newacronym{sc}{sc}{single-cell}
\newacronym{ESC}{ESC}{embryonic stem cells}
\newacronym{MEF}{MEF}{mouse embryonic fibroblasts}
\newacronym{iPSC}{iPSC}{induced pluripotent stem cells}
\newacronym{HSPC}{HSPC}{hematopoietic stem and progenitor cells}
\newacronym{EB}{EB}{embryoid bodies}
\newacronym{PBMC}{PBMC}{peripheral blood mononuclear cells}
\newacronym{HVG}{HVG}{highly variable genes}
\newacronym{DNA}{DNA}{deoxyribonucleic acid}
\newacronym{cDNA}{cDNA}{copy \acrshort{DNA}}
\newacronym{RNA}{RNA}{ribonucleic acid}
\newacronym{mRNA}{mRNA}{messenger \acrshort{RNA}}
\newacronym{FACS}{FACS}{fluorescence activated cell sorting}
\newacronym{ATAC-seq}{ATAC-seq}{transposase-accessible chromatin with sequencing}
\newacronym{RNA-seq}{RNA-seq}{\acrshort{RNA}-sequencing}
\newacronym{4i}{4i}{iterative indirect immunofluorescence imaging}
\newacronym{GEO}{GEO}{gene expression omnibus}
\newacronym{LPS}{LPS}{lipopolysaccharides}
\newacronym{MOA}{MOA}{mode of action}
\newacronym{OHE}{OHE}{one-hot encoding}
\newacronym{MDS}{MDS}{multidimensional scaling}
\newacronym{UMAP}{UMAP}{uniform manifold approximation and projection}
\newacronym{CRISPR}{CRISPR}{clustered regularly interspaced short palindromic repeats}

\printglossary[type=\acronymtype]
